\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2023-01-18 src/tex/ms.tex
Adding comments

There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).

To demonstrate our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_{\theta_1}|h_{\theta_2}\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h_{\theta_1}(f) h^*_{\theta_2}(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the instrinsic parameter point $\theta_1$.
We can then normalize the inner product through
\begin{equation}
    \left[h_{\theta_1}|h_{\theta_2}\right] = \frac{\left(h_{\theta_1}|h_{\theta_2}\right)}{\sqrt{\left(h_{\theta_1}|h_{\theta_1}\right)\left(h_{\theta_2}|h_{\theta_2}\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_{\theta_1}|h_{\theta_2}) \equiv \max_{t_c,\, \phi_c} \left[h_{\theta_1} \, \middle| \, h_{\theta_2} \right]\,
\end{equation}
where $t_c$ and $\phi_c$ are, respectively, the time and phase of coalesence.

@@

@@

@@

\begin{table}[t]
    % \begin{center}
    \centering
    \begin{tabular}{c|c}
    \hline\hline
    $m_1, m_2$ & $\boldsymbol{U}[1.0, 50]\, \mathrm{M}_\odot$ \\ \hline
    $\chi_1, \chi_2$ & [-0.99, 0.99] \\ \hline
    $D$ & [500, 1000]\, Mpc \\ \hline
    $t_c$ & 0.0 \\ \hline
    $\phi_c$ & 0.0 \\ \hline
    Inclincation Angle, $\iota$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Right Ascension, $\delta$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Declination, $\alpha$ & $\boldsymbol{U}[0, 1]$ \\ 
    \hline\hline
    \end{tabular}
    % \end{center}
    \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.}
    \label{tab:priors}
\end{table}

>>>>

\AZ{Add refs here:}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
\AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
\AZ{rephrase this in a more positive fashion:}
NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
In addition, we have restructured the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).

To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the intrinsic parameter point $\theta_1$.
We can then normalize the inner product through
\begin{equation}
    \left[h_{\theta_1}|h_{\theta_2}\right] = \frac{\left(h_{\theta_1}|h_{\theta_2}\right)}{\sqrt{\left(h_{\theta_1}|h_{\theta_1}\right)\left(h_{\theta_2}|h_{\theta_2}\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_{\theta_1}|h_{\theta_2}) \equiv \max_{t_c,\, \phi_c} \left[h_{\theta_1} \, \middle| \, h_{\theta_2} \right]\,
\end{equation}
where $t_c$ and $\phi_c$ are, respectively, the time and phase of coalesence.

@@

\AZ{Broad comments: need to cite which catalog the waveforms came from if any.
The number of waveforms used seems much larger than the number originally used
to calibrate PhenomD. 
If that is true, we can't make a big deal about having a better mismatch compared to
the original version, because of course optimizing over another set will improve 
performance.
Say how many waveforms we use to optimize, how many we compare to, where they came
from.}

@@

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc}

@@

\begin{table}[t]
    % \begin{center}
    \centering
    \begin{tabular}{c|c}
    \hline\hline
    $m_1, m_2$ & $\boldsymbol{U}[1.0, 50]\, \mathrm{M}_\odot$ \\ \hline
    $\chi_1, \chi_2$ & [-0.99, 0.99] \\ \hline
    $D$ & [500, 1000]\, Mpc \\ \hline
    $t_c$ & 0.0 \\ \hline
    $\phi_c$ & 0.0 \\ \hline
    Inclincation Angle, $\iota$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Right Ascension, $\delta$ & $\boldsymbol{U}[0, 1]$ \\ \hline
    Declination, $\alpha$ & $\boldsymbol{U}[0, 1]$ \\ 
    \hline\hline
    \end{tabular}
    % \end{center}
    \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.
    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\cos \delta \in (-1,1)$ }
    }
    \label{tab:priors}
\end{table}
